---
title: "Project 4: Movie Recommender"
author: "Tianhong Yin (tyin7); Deli Yang (deliy2)"
date: "2022-12-11"
output: html_document
---

We wrote code together and Tianhong built the recommendation system.

The link to the movie recommendation app: https://10inxe-tianhong-yin.shinyapps.io/MovieRecommender/

## Exploratory Data Analysis

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library(recommenderlab)
library(DT)
library(data.table)
library(reshape2)
library(Matrix)
library(slam)
```

```{r}
# download and read data
myurl = "https://liangfgithub.github.io/MovieData/"

# ratings data
# use colClasses = 'NULL' to skip columns
ratings = read.csv(paste0(myurl, 'ratings.dat?raw=true'),
                   sep = ':',
                   colClasses = c('integer', 'NULL'),
                   header = FALSE)
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')

# movies data
# in movies.dat, some movie names contain single colon (:), so the method above does not work
movies = readLines(paste0(myurl, 'movies.dat?raw=true'))
movies = strsplit(movies, split = "::", fixed = TRUE, useBytes = TRUE)
movies = matrix(unlist(movies), ncol = 3, byrow = TRUE)
movies = data.frame(movies, stringsAsFactors = FALSE)
colnames(movies) = c('MovieID', 'Title', 'Genres')
movies$MovieID = as.integer(movies$MovieID)
# convert accented characters
movies$Title[73]
movies$Title = iconv(movies$Title, "latin1", "UTF-8")
movies$Title[73]
# extract year
movies$Year = as.numeric(unlist(
  lapply(movies$Title, function(x) substr(x, nchar(x)-4, nchar(x)-1))))

# user data
users = read.csv(paste0(myurl, 'users.dat?raw=true'),
                 sep = ':', header = FALSE)
users = users[, -c(2,4,6,8)] # skip columns
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')

dim(users)
length(unique(ratings$UserID))
```

### Distribution of Ratings

```{r}
# most of the ratings are in the 3-5 range
tmp = data.frame(Rating = 1:5, 
                 freq = as.vector(table(ratings$Rating)/nrow(ratings)))

ggplot(data = tmp, aes(x = Rating, y = freq)) +
  geom_bar(stat="identity", fill = 'steelblue', width = 0.6) + 
  geom_text(aes(label=round(freq, dig=2)), 
                vjust=1.6, color="white", size=3.5) +
  theme_minimal()
```

### Ratings Per User

```{r}
# most users have left less than 500 ratings 
tmp = ratings %>% 
  group_by(UserID) %>% 
  summarize(ratings_per_user = n()) 

summary(tmp$ratings_per_user) # one has 2314 ratings
sort(tmp$ratings_per_user[tmp$ratings_per_user>1300]) # about 9 users have had over 1300 ratings

tmp %>%
  ggplot(aes(ratings_per_user)) +
  geom_bar(fill = "steelblue") + coord_cartesian(c(20, 500))
```
### Ratings Per Movie

```{r}
tmp = ratings %>% 
  group_by(MovieID) %>% 
  summarize(ratings_per_movie = n(), ave_ratings = mean(Rating)) %>%
  inner_join(movies, by = 'MovieID')

summary(tmp$ave_ratings) # the highest average rating is 5, which is the maximum, and the lowest is 1. Mean average rating is 3.239

summary(tmp$ratings_per_movie) # the most popular movie have 3428 ratings and the least popular one has one rating

# The most popular movie is "American Beauty (1999)", which is the first one in the following result, which is followed by the “Star Wars” series. Throughout, "popular" means receiving many ratings. A popular movie may not be a highly-rated movie.
tmp %>% 
  filter(ratings_per_movie > 2000) %>%
  arrange(desc(ratings_per_movie)) %>%
  select(c("Title", "ratings_per_movie"))
```

### Distribution of Genres

```{r}
# create a binary indicator for the 18 genres for each movie
genres = as.data.frame(movies$Genres, stringsAsFactors=FALSE)
tmp = as.data.frame(tstrsplit(genres[,1], '[|]',
                              type.convert=TRUE),
                    stringsAsFactors=FALSE)
genre_list = c("Action", "Adventure", "Animation",
               "Children's", "Comedy", "Crime",
               "Documentary", "Drama", "Fantasy",
               "Film-Noir", "Horror", "Musical",
               "Mystery", "Romance", "Sci-Fi",
               "Thriller", "War", "Western")

length(genre_list)

############## produce the the genre_matrix loaded at the beginning of server.R file
genre_matrix = matrix(0, nrow(movies), length(genre_list))
for(i in 1:nrow(tmp)){
  genre_matrix[i,genre_list %in% tmp[i,]]=1
} 

colnames(genre_matrix) = genre_list
# save(genre_matrix, file = "genre_matrix.rda") # the genre_matrix.rda file for shiny app
####################################################################################

remove("tmp", "genres")

# combine genre_matrix and movie id
genre_com = as.data.frame(genre_matrix)
genre_com = cbind(movies$MovieID, genre_com)
colnames(genre_com) = c("MovieID", genre_list)
genre_com = as_tibble(genre_com)
```

```{r}
# histograms of the 18 genres based on movies: about half of the movies belong to only the drama genre 
data.frame(Genres = genre_list, 
                 Freq = as.vector(colMeans(genre_matrix))) %>% 
  ggplot(aes(reorder(Genres, Freq), Freq, fill = Freq)) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = round(Freq, dig=2)), 
            position = position_stack(vjust = 0.5), 
            color="white", size=3) + 
  coord_flip() + 
  scale_colour_brewer(palette="Set1") + 
  labs(y = 'Frequency', x = 'Genre')

# histograms of the 18 genres based on popularity, which measures the number of ratings a movie receives
tmp = ratings %>% 
  left_join(data.frame(MovieID = movies$MovieID, genre_matrix), 
            by = "MovieID") %>%
  select(-c("UserID", "MovieID", "Rating", "Timestamp"))

data.frame(Genres = genre_list, 
                 Popularity = as.vector(colMeans(tmp))) %>% 
  ggplot(aes(reorder(Genres, Popularity), Popularity, fill = Popularity)) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = round(Popularity, dig=3)), 
            position = position_stack(vjust = 0.5), 
            color="white", size=3) + 
  coord_flip() + 
  labs(y = 'Popularity', x = 'Genre')

tmp = rowSums(genre_matrix)
summary(tmp)

movies[which(tmp==6), ] # Movies can belong to up to six genres in the data, which is "The Transformers" (1986) that is categorized into six genres: Action, Animation, Children’s, Sci-Fi, Thriller, and War

# range(movies$Year)% from 1919 to 2000
tmp = data.frame(Year = movies$Year, genre_matrix) %>%
  group_by(Year) %>%
  summarise_all(sum)

tmp[,-1] = apply(tmp[, -1], 2, cumsum)
tmp[,-1] = tmp[,-1]/sum(tmp[nrow(tmp), -1])
print(round(tmp[nrow(tmp),-1], dig=3)) # the cumulative percentages of the 18 genres that users can access from 1919 to 2000, which is stored in a 81-by-19 matrix tmp

# A graphical display of the 18 genres
tmp = reshape2::melt(tmp, id.vars="Year") 
tmp %>%
  ggplot(aes(Year, value, group = variable)) +
  geom_area(aes(fill = variable)) + 
  geom_line(aes(group = variable), position = "stack")
```

## System I: Recommendation Based on Genres

### First Scheme: Recommend 12 Top-Rated Movies (Applied in Shiny App)

* Note: "top-rated" is based on users' average rating (on a 1-5 scale).

```{r}
# suppose we know the user's favorite genre is Comedy
genre_user = "Comedy"

# find out the genre 
genre_movies = genre_com[(which(genre_com[[genre_user]]== 1)),]

# reference: exploratory data analysis I
movie_rating = ratings %>% 
  group_by(MovieID) %>%
  summarize(ratings_per_movie=n(), ave_ratings = mean(Rating)) %>%
  inner_join(genre_movies, by = 'MovieID') %>%
  inner_join(movies, by = 'MovieID')

# sort by average ratings (descending order)
scheme1 = movie_rating[order(-movie_rating$ave_ratings),]

# select the first 12 rows
top_12 = scheme1[1:12, ]

# only display certain columns
top_12[ , c("MovieID", "Title", "ave_ratings", "Comedy")]
```

The following function also works, which is used in building the recommendation system app. We just need to fill in a genre in the function and we can get the 12 top-rated movies to recommend.

```{r}
# average rating for each movie in a specific genre
allmovies_genre = function(genre) {
  movieid_genre = genre_matrix[,genre] == 1
  movies_genre = movies[movieid_genre,]
  
  ratings <- ratings %>% 
    group_by(MovieID) %>% 
    mutate(avg_ratings = round(mean(Rating), dig=2)) %>%
    select(-Rating) 
  
  avg_rating<- distinct(ratings, MovieID, avg_ratings)
  
  ret = movies_genre %>%
    left_join(avg_rating, by = 'MovieID') %>% 
    replace(is.na(.), 0) %>% 
    mutate(avg_rating = dense_rank(desc(avg_ratings))) %>% 
    arrange(desc(avg_ratings)) %>%
    distinct()
  return(ret)
}

# top N rated movies based on the average rating
topN_movies_genre = function(top_n, genre) {
  full_movies_by_genre = allmovies_genre(genre)
  ret = full_movies_by_genre %>%
    arrange(desc(avg_ratings)) %>%
    head(top_n) %>%
    select('MovieID', 'Title','avg_ratings', 'Genres')
  return(ret)
}

# a user choosing "Comedy" produces the same results:
topN_movies_genre(12, "Comedy")
```


### Second Scheme: Recommend Top 12 Most Popular Movies

* Note: "popularity" is measured by the number of users who give ratings to the movies. A larger number of ratings indicates more popular.

```{r}
#### suppose we know the user's favorite genre is Comedy, we can use the previous results:
# genre_user = "Comedy"
# 
# # find out the genre 
# genre_movies = genre_com[(which(genre_com[[genre_user]]== 1)),]
# 
# # reference: exploratory data analysis I
# movie_rating = ratings %>% 
#   group_by(MovieID) %>%
#   summarize(ratings_per_movie=n(), ave_ratings = mean(Rating)) %>%
#   inner_join(genre_movies, by = 'MovieID') %>%
#   inner_join(movies, by = 'MovieID')

# sort by the number of ratings (descending order)
scheme2 = movie_rating[order(-movie_rating$ratings_per_movie),]

# select the first 12 rows
top_12 = scheme2[1:12, ]

# only display certain columns
top_12[ , c("MovieID", "Title", "ratings_per_movie", "Comedy")]
```

Also works in the following function:

```{r}
allmovies_pop = function(genre, top_n) {
  movieid_genre = genre_matrix[,genre] == 1
  movies_genre = movies[movieid_genre,]
  
  ratings <- ratings %>% 
    group_by(MovieID) %>% 
    summarize(ratings_per_movie=n()) 
  
  ret = merge(ratings, movies_genre) %>%
    filter(grepl(genre_user, Genres)) %>%
    replace(is.na(.), 0) %>% 
    arrange(desc(ratings_per_movie)) %>%
    select('MovieID', 'Title', 'ratings_per_movie','Genres') %>%
    head(top_n)
  
  return(ret)
}

allmovies_pop("Comedy", 12) # the same results
```

## System II: Collaborative Recommendation System - UBCF and IBCF

### UBCF
```{r}
library(recommenderlab)
i = paste0('u', ratings$UserID)
j = paste0('m', ratings$MovieID)
x = ratings$Rating
tmp = data.frame(i, j, x, stringsAsFactors = T)
Rmat = sparseMatrix(as.integer(tmp$i), as.integer(tmp$j), x = tmp$x)
rownames(Rmat) = levels(tmp$i)
colnames(Rmat) = levels(tmp$j)
Rmat = new('realRatingMatrix', data = Rmat)

# normalization
train = Rmat[1:500, ]
test = Rmat[501, ]

testmat= as(test, "matrix")
data = as(train, "matrix")
user.means = rowMeans(data, na.rm = TRUE)
data = data - user.means

newdata = as(Rmat[501, ], "matrix")
newuser.mean = mean(newdata, na.rm = TRUE)
newdata = newdata - newuser.mean

# calculate the similarity
sim = rep(0, dim(data)[1])
for(i in 1:length(sim)) {
  tmp.y = as.vector(newdata)
  ind.y = which(!is.na(tmp.y))
  tmp.x = data[i, ]
  ind.x = which(!is.na(tmp.x))
  ind = intersect(ind.x, ind.y)
  if (length(ind) > 0){
    tmp.x = tmp.x[ind]
    tmp.y = tmp.y[ind]
    sim[i] = sum(tmp.x * tmp.y) / sqrt(sum(tmp.x^2) * sum(tmp.y^2))
  }
}
sim.ubcf = (1 + sim)/2

# find the top 20 similarity score and the user position in data
rank.ubcf = rank(-sim.ubcf)
top20id.ubcf = match(c(1:20), rank.ubcf)

# prediction
data_0_1 = data
data_0_1[!is.na(data_0_1)] <-1
data_0_1[is.na(data_0_1)] <-0
data[is.na(data)] <-0

mypred=vector()
for (i in 1:ncol(data)) {
  num = sum(sim.ubcf[top20id.ubcf]%*%data[top20id.ubcf,i])
  deno = sum(sim.ubcf[top20id.ubcf]%*%data_0_1[top20id.ubcf,i])
  mypred[i] = (num/deno) + mean(testmat, na.rm = TRUE)
}
naposi = which(!is.na(testmat))
mypred[naposi] <- NA

# recommenderlab
recommender.UBCF <- Recommender(train, method = "UBCF",
                                parameter = list(normalize = 'center', 
                                                 method = 'Cosine', 
                                                 nn = 20))

p.UBCF <- predict(recommender.UBCF, test, type="ratings")
p.UBCF <- as.numeric(as(p.UBCF, "matrix"))

sum(is.na(p.UBCF) != is.na(mypred)) ### zero
max(abs(p.UBCF - mypred), na.rm = TRUE)  ### less than 1e-06 
```

### IBCF (Applied in Shiny App)
```{r}
# Use the previous train and test data
# normalization
train = Rmat[1:500, ]
test = Rmat[501, ]

data = as(train, "matrix")
user.means = rowMeans(data, na.rm = TRUE)
data = data - user.means
newdata = as(Rmat[501, ], "matrix")

# similarity
item.sim = proxy::simil(data, by_rows = FALSE, method = "cosine")
item.sim_mat = as.matrix((1+item.sim)/2)

for (i in 1:dim(item.sim_mat)[1]){
  top30id.ibcf = tail(order(item.sim_mat[i,], decreasing = FALSE, na.last = FALSE), 30)
  item.sim_mat[i,-top30id.ibcf] =NA
}

# prediction
ra = as.vector(newdata)
ra_mat = matrix(rep(ra, each=3706), nrow=3706)
dat = item.sim_mat*ra_mat
naposi = which(is.na(newdata))
mypred2 = rep(NA, times=ncol(newdata))

for (i in 1:length(naposi)) {
  num = sum(dat[naposi[i],],na.rm = TRUE)
  deno = sum(!is.na(dat[naposi[i],]))
  mypred2[naposi[i]] = num/deno
}

mypred2[is.infinite(mypred2)] = NA
mypred2[is.nan(mypred2)] =NA

recommender.IBCF <- Recommender(train, method = "IBCF",
                                parameter = list(normalize = 'center', 
                                                 method = 'Cosine', 
                                                 k = 30))

p.IBCF <- predict(recommender.IBCF, test, type="ratings")
p.IBCF <- as.numeric(as(p.IBCF, "matrix"))

## first output: should be less than 10
sum(is.na(p.IBCF) != is.na(mypred2))  

## second output: should be less than 10%
mydiff2 = abs(p.IBCF - mypred2)
sum(mydiff2[!is.na(mydiff2)] > 1e-6) / sum(!is.na(mydiff2)) 
```

